\documentclass{article}

% Language setting
% Replace `english' with e.g. `spanish' to change the document language
\usepackage[english]{babel}

% Set page size and margins
% Replace `letterpaper' with `a4paper' for UK/EU standard size
\usepackage[letterpaper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{upgreek}
\usepackage{amsthm}

\title{Set Theory and Logic Notes}
\author{Nat}

\newcommand{\defn}{\textbf{Def: }}
\newcommand{\dfn}{\textbf{Def: }}

\newcommand{\set}[1]{\{#1\}}


\begin{document}
    \maketitle

    \section{Introduction}

% Group theory. Group is a set of objects .

% G - a set

% . - binary function

% -1 - inverse function

% 1 - identity

% Axioms of group theory:

% For all a,b,c in G

% associative - a.b. . c =

    We can abbreviate statements, e.g. $\neg (x=y)$ and $x \neq y$
    there exists a unique x satisfying $P(x) \rightarrow \exists! x P(x)$
    expansion: $\exists x (P(x) \and \forall y (P(y) \rightarrow y=x))$
    note we can give a tree which shows how a formula is constructed.
    The connectives and quanitifers are the internal nodes and root, terminals are predicates/relations.

    Connectives are functions on truth values, can be described using truth tables.
    Examples:

    \begin{tabular}{c|c}
        P  &  $\neg$ P\\
        \hline
        T  & F \\
        F & T
    \end{tabular}

    \vspace{1em}

    \begin{tabular}{c|c|c}
        P & Q & P $\wedge$ Q\\
        \hline
        T & T & T \\
        T & F & F \\
        F & T & F \\
        F & F & F \\
    \end{tabular}

    \vspace{1em}

    \begin{tabular}{c|c|c}
        P & Q & P $\implies$ Q\\
        \hline
        T & T & T \\
        T & F & F \\
        F & T & T \\
        F & F & T \\
    \end{tabular}

    \vspace{1em}
    \begin{tabular}{c|c|c}
        P & Q & P $\iff$ Q\\
        \hline
        T & T & T \\
        T & F & F \\
        F & T & F \\
        F & F & T \\
    \end{tabular}

    \vspace{1em}
    \textbf{Logical Equivalence}
    \begin{itemize}
        \item Two statements are logically equivalent if for any truth assignments of truth values to the predicates, both statements have the same resulting truth value. $s_1 \equiv s_2$
    \end{itemize}

    \emph{Ex}: $P \implies Q$ is logically equivalent to $(\neg P ) \vee Q$

    \begin{tabular}{c|c|c}
        P & Q & $(\neg P) \vee Q$\\
        \hline
        T & T & T \\
        T & F & F \\
        F & T & T \\
        F & F & T \\
    \end{tabular}
    \vspace{1em}

    \textbf{Def:} A statement is \emph{valid} if it is true for any truth assignment to the predicates
    \emph{Ex:} $P \vee \neg P$


    \section{Set Theory}
    \begin{itemize}
        \item First to study this was Cantor
        \item \"size\" of infinite sets
        \item two sets have the same size if we can put a bijection between the two
        \item properties of sets.
    \end{itemize}

    \underline{Idea:} Sets are determined by their members. $\{1,2,3,4\} \equiv \{2,3,1,4\}$

    \underline{Goals}
    \begin{itemize}
        \item Formalize Sets
        \item Introduce Axioms of Zermelo-Frankel Set Theory with choice (ZFC)
        \item Cardinality
        \item represent standard mathematical objects as sets
        \item ZFC as a foundation for all of mathematics
    \end{itemize}

    \underline{Note}
    \begin{itemize}
        \item Everything will be a set
        \item Language of set theory will be built from two predicates!
        \begin{itemize}
            \item = : equality
            \item $\in$ : set membership. $x\in y$ will be true when x is an element of y
        \end{itemize}
    \end{itemize}

    \emph{Ex:} $\emptyset$ , empty set. $\{\emptyset\} \neq \emptyset$. Then we can build natural numbers. 2: $\{\emptyset,\{\emptyset\}\}$

    \vspace{1em}
    \noindent\textbf{\emph{The Axioms of ZFC}}

    \underline{Language} built from =, $\in$
    \begin{itemize}
        \item Axiom 0: \emph{set existence}, There is a set. $\exists x (x=x)$. (axiom 0 will be redundant later).
        \item Axiom 1: \emph{extensionality}, Sets are determined by their elements
        \item Axiom 2: \emph{Foundation} There are no infinite descending chains of sets with respect to elementhood. i.e. No $..... \in x_2 \in x_1 \in x_0$. implies no element is an element of itself
        \item Axiom 3: \emph{comprehension scheme}
        \begin{itemize}
            \item For every formula $\varphi(x)$ and every set $Z$ we can form the set $\{x \in Z : \varphi(x)\}$.
            Those x which are elements of $Z$ which satisfy $\varphi(x)$
            \item one axiom for each formula $\varphi(x)$
        \end{itemize}
        \item Axiom 4: \emph{pairing}, Given sets $x$ and $y$, there is a set containing both $x$ and $y$
        \item Axiom 5: \emph{union}, Given a family of sets, $F$, there is a set containing the union of all of the sets in $F$
        \item Axiom 6: \emph{replacement scheme}, Given a formula $\varphi$ which defines a function $f$ and a set $Z$, there is a set which consists of the range of $f$ applied to $Z$
        \item Axiom 7: \emph{infinity} (makes 0 redundant), there is an infinite set
        \item Axiom 8: \emph{power set}, given a set $X$ there is a set $\mathcal{P}(X)$ which contains as elements all of the subsets of $X$
        \item Axiom 9: \emph{axiom of choice (AC)}, given a collection of disjoint non-empty sets, there is a set which contains exactly one element from each set in the collection.
        \begin{itemize}
            \item \emph{Ex}: $\mathbb{R}$ $x~y$ if $x-y \in \mathbb{Q}$

            $[x] = \{y: y~x\}$

            $\mathcal{e}=\{[x]: x \in \mathbb{R} \}$

            if x not ~ y then [x] and [y] are disjoint$ [x] \neq \emptyset$
            \item AC says there is a set containing exactly one element of each equivalence set
            \item this is called a \emph{Vitali} set.
        \end{itemize}

        \underline{Note:}
        \begin{itemize}
            \item ZFC denotes Axioms 1-8
            \item ZF denotes Axioms 1-8
            \item ZC and Z denote ZFC and ZF with replacement scheme removed
            \item Z-, ZF-, ZC-, ZFC- denote removing foundation axiom
            \item \"most\" standard mathematics can be don in ZC-
        \end{itemize}

    \end{itemize}


    \subsection{Axiom 0 - Set Existence}
    \subsection{Axiom 1 - Extensionality}
    \subsection{Axiom 2 - Foundations}
    \subsection{Axiom 3 - Comprehension Scheme}
    \subsection{Axiom 4 - Pairing}

    \textbf{Recall} - Ordered Pair: $<x,y> = \{\{x,y\},\{x\}\}$

    \subsection{Axiom 5 - Union}
    \[\forall \mathcal{F} \exists A \forall Y \forall x ((x \in Y \wedge Y \in mathcal{F}) \implies x \in A
    \]
    \underline{Ex} $Y_1, Y_2, \dots, Y_k \in \mathcal{F}$

    $A \supseteq Y_1 \cup Y_2 \dots \bigcup Y_k$

    \vspace{1em}
    \underline{Note}: Comprehension allows us to form the et whose elements are exactly those x so that $x \in Y$ for some $Y \in \mathcal{F}$. Denote this set $\bigcup \mathcal{F}$, i.e. $Y^\cup \in \mathcal{F}^Y$

    \underline{Ex:} $x \cup y = \bigcup \{x,y\} =\{z : z \in x \cup z \in y\}$

    $x \cup y \cup z = \big \{x,y,z\}$

    \vspace{1em}
    \underline{Ex:} $\bigcup\{x\} = \{z : z \in x\} = x$

    $ \bigcup\{\{x\}\} = \{x\}$ and $\bigcup \bigcup\{\{x\}\} = x$

    \vspace{1em}
    \underline{Def}: When $\mathcal{F} \neq \emptyset$ we can define intersection. $\bigcap \mathcal{F} = \bigcup_{Y \in \mathcal{F}} Y$

    $=\{x : \forall Y \in \mathcal{F} (x \in y)\}$ (naive)

    $\{x\in Y_0 : \forall Y \in \mathcal{F}(x \in y)\}$ (proper use of comprehension.)

    where $Y_0$ is some element of $\mathcal{F}$

    \underline{Note} If $\mathcal{F} = \emptyset$ then $\bigcap \mathcal{F} would be a universal set$

    \vspace{1em}
    \underline{Def:} The \emph{ordinal successor function}

    $S(x) = x \cup \{x\}$

    \underline{Ex:} $S(\emptyset) = \emptyset \cup \{\emptyset\} = \{\emptyset\}$ and $S(\{\emptyset\}) = \{\emptyset\} \cup \{\{\emptyset\}\} = \{\emptyset,\{\emptyset\}\}$

    $0 = \emptyset$

    $1 = S(0)$

    $2 = S(1) = S(S(0))$

    $n+1 = S(n)$

    \underline{Note} $n = \{ k : k < n\}$, n has exactly n elements. $n \leq m iff n \subseteq m$ and $n < m iff n \in m$

    \underline{Informally} a natural number is a set obtained by some finite number of iteration of the successor applied to $\emptyset$


    \subsection{Relations and Functions}

    \underline{Def} R is a (binary) relation if R is a set of ordered pairs, i.e. $\forall z \in R \exists x \exists y z = <x,y>$

    \underline{Note} If $\phi(z) is a formula the abbreviation \forall z \in R \phi(z) and \exists z \in R \phi(z)$ means

    $\forall Z \in R \phi(z) means \forall z (z \in R \implies \phi(z))$

    $\exists z \in R \phi(z) means \exists z (z \in R \and \phi(z))$

    \vspace{1em}
    \underline{Def} $dom(R) =^* \{x : \exists y <x,y> \in R\}$

    $ran(R) =^* \{y: \exists x <x,y> \in R\}$

    * need to see that there is a set $Z$ so that all $x$ and $y$ needed in these definitions are elements of $Z$

    If $<x,y> \in R$

    $\{\{x\},\{x,y\}\}$

    $\{x\} \in \bigcup R$ and $\{x,y\} \in \bigcup R$ also
    $x \in \bigcup \bigcup R$ and $x,y \in \bigcup \bigcup R$

    So $dom(R) = \{x \in \bigcup \bigcup R : \exists y <x,y> \in R\}$
    $ran(R) = \{y \in \bigcup \bigcup R : \exists x <x,y> \in R\}$

    \vspace{1em}
    \underline{Def} Properties of Relations

    Let $R$ be a relation and $A$ some set. We will say that $R$ is\ldots
    \begin{itemize}
        \item \emph{transitive on A} if $\forall x,y,z \in A$
        \[((x R y \and y R z) \implies x R z\]
        where $x R y$ means $<x,y> \in R$
        \item \emph{reflexive on A} if $\forall x \in A (x R x)$
        \item \emph{irreflexive on A} if $\forall x \in A \neg(x R x)$
        \item \emph{symmetric on A} if $\forall x,y \in A (xRy \leftrightarrow y R x)$
        \item \emph{partially orders A strictly} if R is irreflexive on A and transitive on A
        \item \emph{satisfies trichotomy on A} if  $\forall x,y \in A (x R y \lor y R x \lor x=y)$
        \item \emph{totally orders A strictly} if R partially orders A strictly and R satisfies trichotomy on A
        \item \emph{equivalence relation on A} if R is reflexive on A, symmetric on A, and transitive on A.
    \end{itemize}

    \underline{Ex} $A = \mathbb{N}$ and  $x R y$ if $x \equiv_3 y$. i.e. x-y is a multiple of 3. R is an equivalence relation on A.

    e.g. $x R y$ and $y R z$ then $x-y =3j$ for  some j and $y-z = 3k$ for some k
    $x-z=(x-y) + (y-z) = 3j + 3k = 3(j+k)$ so $x R z$

    \underline{Ex} $A = \mathbb{N}$  $x R y$ if x divides y is reflexive on A and transitive on A.

    \underline{Ex} $A = \mathbb{N}$ $x R y$ if $x < y$ R totally order A strictly

    \underline{Def} Let R be a relation and A a set. The restriction of R to A, $R \upharpoonright A = \{<x,y> \in R : x \in A \}$


    \underline{Def}
    \begin{itemize}
        \item R is a \emph{function} if R is a relation $\forall x \in dom(R) \exists! y <x,y> \in R$
        \item F, $x \in dome(R)$ we will write R(x) to mean the unique y with $<x,y> \in R$
    \end{itemize}

    \underline{Ex} $F = \{<0,0> ,<1,0>,<2,2>\}$ is a function $dom(F) = \{0,1,2\}$ and $ran(F) = \{0,2\}$ $F(0) = 0, F(1) = 0, F(2) = 2$

    \underline{Def} $F : A \rightarrow B$ means F is a function $dom(F) = A$ and $ran(F) \subseteq B$
    F above $F: \{0,1,2\} \rightarrow \{0,1,2\}$
    $F : A \rightarrow_{onto} B$ or $F: A \twoheadrightarrow B$
    if $F: A \rightarrow B$ and $ran(F) = B$ i.e. F is surjective or onto.
    $F: A \rightarrow^1-1 B$ or $F : A \hookrightarrow B$ if $F: A \rightarrow B$ and $\forall x_1, x_2 \in A$ then $(F(x_1) = F(x_2) \implies x_1=x_2)$ i.e. F is injective or one-to-one.
    $F: A \rightarrow^{1-1}_{onto} B$ or $F : A \leftrightarrow B$ as F is surjective and injective

    \underline{Ex}: $F(x) \sin X$
    $F : \mathbb{R} \rightarrow \mathbb{R}$
    $F : $

    \underline{Ex} $\sin([0,\pi / 2]) = [0,1]$ but we need to be more careful because everything is a set.

    \underline{Ex:} $F=\{<0,0>,<1,0>,<2,2>\}$ Recall $2 = \{0,1\}$
    $F(2)=2$
    the range of F applied to the set 2 is $\{F(0), F(1)\}=\{0,0\}=\{0\}$

    \underline{Def:} For a function F and a set A. $F" A$ (F applied to A) is $F" A = ran(F\upharpoonright A)=\{F(x) : x \in A \}$
    \underline{Ex} above F(2)=2, $F"2=\{0\}$


    \underline{\textbf{\emph{Jan 28}}}

    \emph{Note} axioms we have so far don't allow us to build many functions or relations.
    Given set A,B a function f: a $\rightarrow$ b will be a subset of the direct product A X B but axioms so far don't allow us to construct A X B
    even $\{\emptyset\}$ X B can't be produced

    \underline{Ex} Suppos ewe had the set of natural numbers $\mathbb{N}$
    And we want ot define $F(n) = S(n)$ (successor function).
    we need the set $\{<n,S(n)> : n \in \mathbb{N}\}\}$
    Let $\varphi(x,y) : y=S(x)$ i.e. $y = x \cup \{x\}$
    woule like to see the \"range\" of $\varphi$ exists applied to $\mathbb{N}$

    \subsection{Axiom 6 - Replacement Scheme}
    For every formula $\varphi(x,y)$ so that the variable $B$ does not occur in $\varphi$ we have the axiom.
    \[ \forall A ((\forall x \in A \exists! y \varphi(x,y)) \implies (\exists B \forall x \in A \exists y \in B \varphi(x,y)))\]

    \emph{Note} called replacement since we replace each x in A by the unique y with $\varphi(x,y)$
    Comprehension gives us the set range (f/A) where f is the function defined by $\varphi$

    \underline{Def: Direct Product} (Naive)
    \[S \times T = \{x : \exists s \in S \exists t \in T x=<s,t>\}\]
    need to see there is a set containing all such x's
    \underline{Lem:} Given set S and T there is a set Z so that for every $s \in S$ and $t \in T <s,t> \in Z$
    \underline{Cor:} THe direct prodcut exists: $S \times T = \{x \in Z : \exists s \in S \exists t \in T x=<s,t>\}$

    \underline{\emph{Proof of Lem:}}
    \begin{itemize}
        \item Fix $s \in S$.
        \item Let $\varphi(x,y)$ be $y=<s,x>$
        \item Then $\{s\} \times T = \{<s,t> : t \in T\} = \{<w \in B : \exists t \in T w = <s,t>\}$ where B is obtained from replacement applied to the set $A = T$ and $\varphi(x,y)$
        \item Next Let $\varphi(x,y)$ be $y = \{s\} \times T$ and $A=S$
        \item By replacement the set $n = \{\{s\} \times T : s \in S\}$
        \item Then $\bigcup n = \bigcup_{s \in S}\{s\} \times T = S \times T$
    \end{itemize}

    \textbf{Note:} We can now also define functions by formulas.

    \underline{Lemma:} Suppose $\forall x \in A \exists! y \varphi(x,y)$ then there is a function $f$ with $dom(f)=A$ so that for every $x \in A$ $f(x)$ is equal to the unique y so that $\varphi(x,y)$

    \emph{Pf:} Let $B$ be the set given by Replacement.
    Then $f = \{w \in A \times B : \exists x \exists y (w=<x,y>) \wedge \varphi(x,y))\}$

    \underline{Ex:} $A = \mathbb{N} \varphi(x,y) is y=S(x)$ then there is a function $f$ so with $dom(f) = \mathbb{N}$ and $f(n) =S(n)$ for all $n \in \mathbb{N}$

    \textbf{Note:} Multivariable function are relations can be defined as well

    \underline{Ex:} $f(x,y)$ can be viewed as $f: X \times Y \rightarrow Z$

    \textbf{Def:} Given a relation $R$ the \underline{inverse relation} $R^{-1}=\{<y,x> \in ran(R) \times dom(R) : <x,y> \in R\}$

    \underline{Note:} If f is a function, $f^{-1}$ won't necessarily be a function unless f is one-to-one

    \textbf{Def:} Given relations $F$ and $G$ the \underline{composition} $G \circ F$ is given by $G \circ F = \{<x,z> \in dom(F) \times ran(G): \exists y (<x,y> \in F \wedge <y,z> \in G)\}$
    \underline{Note:} If $F$ and $G$ are functions and $ran(F) \subseteq dom(G)$ then $G \circ F$ is the usual composition with $(G \circ F)(x)=G(F(x))$

    \underline{Def:} Suppose $(S,<) and (T,\prec)$ are relations, i.e. < is a relation with $dom(<)$ and $ran(<)$ subset of $S$ and similarly for $(T,\prec)$.
    Then the \underline{lexicographic product} on $S \times T$ is defined by relation $<_{lex}$ given by $<s,t>  <_{lex} <s',t'>$ when $s < s' \vee (s=s' \wedge t \prec t')$

    \underline{Ex:} $S=F=\{0,1,2\}$ < and $\prec$ be the usual ordering $0<1<2, 0 \prec 1 \prec 2$
    $<0,0> <_{lex} <0,1> <_{lex} <0,2> <_{lex} <1,0> \dots <_{lex} <2,2>$

    \emph{Note:} (homework) If < and $\prec$ are strict total orderings then so is $<_{lex}$.

    \underline{Def:} let $R$ be a relation on the $A$  and $S$ a relation on $B$.
    A function $F$ is an \underline{isomorphism} from $(A,R)$ onto $(B,S)$.
    if $F: A \rightarrow B$ so that $\forall x,y \in A$ $x R y \iff F(x) S F(y)$
    we say $(A,R)$ is isomorphic to $(B,S)$, $(A,R,) \cong (B,S)$ if there is a function $F$ which is an isomorphism of $(A,R)$ into $(B,S)$

    \underline{Def:} Let $R$ be an eqiuvalence relation on $A$ (domain and range of R are subsets of A).
    For $x \in A$ let the \underline{equivalence class} of $x$ be $[x] = \{y \in A : y R x\}$
    \emph{Note } $x R y \iff [x] = [y]$

    \underline{Def: } $A/R = \{[x] : x \in A\}$
    i.e. $\{w \in B : \exists x \in A w \ [x]\}$
    where $B$ is the set produced by applying Replacement scheme to the formula $\varphi(x,y) : y=[x]$

    \subsection{Interlude}
    Finite models of parts of Set Theory
    \emph{Note:} can describe a structure on a finite domain where we interpret $\in$ explicitly

    \textbf{Ex:} $X = \{a,b,c\}$ and $a \in b, b \in c$
    i.e. $\in$ is interpreted by relation $E$ where $E = \{ <a,b> , <b,c>\}$

    we can draw a directed graph where nodes are elements of x and arrow represent ordered pairs in $E$.

    Given such a structure, which axioms are satisfied when we interpret $\in$ by $E$

    \underline{Extensionality} two sets are the same exactly if they have the same elements:
    a has no elements, b has a single element a, and c has a single element b.
    satisfies extensionality as they are seperate.

    \underline{Pairing} given two sets then a set exists containing both:
    not satisfied - no set has both a and b as elements.

    \underline{Foundation} no infinite chain of membership:
    satisfied

    \emph{Ex:} $X = \{a,b,c\}$
    $E = \{<a,b>,<b,c>,<c,a>\}$
    (cyclic triangle)

    \underline{Extensionality}: a has element c, b has a, and c has b, satisfying
    \underline{Pairing} fails
    \underline{Foundation} fails

    \emph{Ex} $E =\{<a,c>,<b,c>\}$ (binary tree)
    Foundation and Pairing Fails, and so does Extensionality as a nd b are both empty sets.

    \emph{Ex} $X=\{a\}$ and $E=\{<a,a>\}$ (self-loop)
    Foundation fails, extensionality is trivially satisfied as there is only one set, and it satisfies pairing.

   \emph{\underline{\textbf{Jan 30}}}

    recall let r be a relation
    r is a strict toal order on A if - R is transitive on A, irreflixive on A and satisfies trichotomy
    \textbf{Def}: Let R be a relation. We say $a \in X$ is R-minimal in X if $\neg \exists z ( z \in X \wedge Z R a)$

    \emph{Note:}    a is not necessarily least, i.e. there may be more than one element of x which is R-minimal in X.

    \underline{EX:} $x=\mathbb{N}$ $n R m$ when $n<m$ 0 is the unique \underline{R-minimal} element in x.

    Let $A \subseteq \mathbb{N}$ $A \neq \emptyset$ then A has an R-minimal in A.

    \underline{Ex:} $x = \{ n \in \mathbb{N} : n \geq 2\} n R m$ when m is a proper multiple of n.
    p is R-minimal in X exactly when p is prime.
    infinite R-minimal

    \underline{Ex:} $x=\mathbb{Z}$ $n R m$ when $n < m$ No R-minimal element in X.
    no R-minimal

    \emph{Note:} we can also define R-maximal

    \textbf{Def} A relation R is \underline{well-founded} on X if for every non-empty $A \subseteq X$ there is an $a \in A$ which is R-minimal in A

    \underline{Ex:} $x=\mathbb{N}$ $n R m$ when $n < m$ R is well-founded on X.

    \underline{Ex:} $x = \{n \in \mathbb{N} : n \geq 2\}$ $n R m$ when m is a propers multiple of n.
    R is well-founded on X.
    Let A be a non-empty subset of X\@.
    pick $a_0 \in A$
    if $a_0$ is R-minimal in A we are done (no integer (element in A) is a proper divisor of it)
    else if not there is $a_1 \in A$ so that $a_1$ is a proper divisor of $a_0$
    Note $a_1 < a_0$
    if $a_1$ R-minimal in A we are done \dots etcetera
    This must stop at some point as we can't go below two since the sequence would be decreasing.
    The last term will be R-minimal in A.

    \underline{Ex:} $x \in \mathbb{Z}$ $n R m$ when $n < m$ R is not well-founded
    $A = \mathbb{Z}$ has no R-minimal element

    \underline{Ex:} $x=[0,1]$ $x R y$ when $x < y$.
    R is not well-founded on X.
    $A = (0,1)$ has no R-minimal element because if $a \in A$ $a>0$ so let $b=a/2$ and $b<a$ showing we can find something smaller.

    \textbf{Def:} A relation R is \underline{well-orders} a set A if R totally orders A strictly (transitive and satisfies trichotomy) and R is well-founded on A.

    \underline{Ex:} < well-orders $\mathbb{N}$.

    \underline{Ex:} < does not well-order $\mathbb{Z}$ or $[0,1]$

    \underline{Ex:} proper multiple relation does not well-order $\{n \in \mathbb{N} : n \geq 2\}$

    \textbf{Def}: A \underline{well-ordering} is a pair $(A ; R)$ where A is a set and R is a relation which well-orders R.

    \underline{Ex:} $(\mathbb{N};<)$ is a well-ordering

    \emph{Note:} For a well-ordering $(A;R)$ every non-empty $X \subseteq A $ has an \underline{R-least} element, i.e. there is a unique element of X which is R-minimal in X.
    suppose both x and y are R-minimal in X
    either x=y or $x R y$ meaining y is not R-minmal or vice versa.

    \emph{Note:} If $(S;<)$ and $(T;\prec)$ are both well-orderings then so is their lexicographic product.

    \emph{Note:} If R well-orders A and $B \subseteq A$ is a non-empty subset then R well-orders B.

    \underline{ex:} $(N;<)$

    B = primes

    < is a well-ordering of B
    
    \underline{Ex:} $0=\emptyset$ , $1 = \{\emptyset\}$, $2 = \{\{\emptyset\}\}\}$
    
    $x = \{0,1,2\}$ and $R = \in$
    
    Then $\in$ well-orders X.
    
    \emph{Note:} If X is a finite set then any strict total ordering of X is actually a well-ordering of X.
    
    
    \textbf{\underline{Ordinals}}
    Recall: $0=\emptyset$ and $n+1=S(n)$ where $S(x) = x \cup \{x\}$
    Informally: Natural numbers are 0, S(0), S(S(0))
    
    \textbf{Def}: A set $Z$ is \underline{transitive} if $\forall y \in Z (y \subseteq Z)$
    i.e. $\forall x \forall y ((x \in y \wedge y \in Z) \rightarrow x \in z)$
    
    \underline{Ex:} $z=2=\{\emptyset\{\emptyset\}\}$
    $y=\emptyset$
    $\emptyset \subseteq Z$
    $y=\{\emptyset\}$
    $\{\emptyset\} \subseteq Z$ since $\emptyset \in Z$
    so Z is transitive
    
    \underline{Ex:} $Z=\{\{\emptyset\}\}$ is not transitive
    $y=\{\emptyset\}$
    $y \in Z$
    $x=\emptyset$
    $x \in y$
    but $\emptyset \notin Z$
    
    \emph{Note:} The $\in $ relation behaves transitively with respect to the set Z.
    Z is transitive of $(x \in y \wedge y \in z) \implies x \in z$ for any x and y.

    (natural numbers are transitive)
    
    \textbf{Def} Z is a (von Neumann) \underline{ordinal} if Z is transitive, and Z is well-ordered by $\in$
    
    \underline{Ex:} $2 = \{\emptyset\{\emptyset\}\}$ is an ordinal
    elements: $\emptyset, \{\emptyset\}$
    
    $3=\{0,1,2\}$ is an ordinal
    
    every natural number is an ordinal (see this ?later on?)
    
    \emph{Note: } we usually use Greek letters to denote ordinal.
    
    \textbf{Def} For ordinal $\alpha$ and $\beta$:
    
    $\alpha < \beta$ means $\alpha \in \beta$
    
    $\alpha \leq \beta$ means $\alpha \in \beta$ or $\alpha = \beta$
    
    \textbf{Def} ON or Ord is the class of all ordinals.
    
    \emph{Note:} $ON$ is not a set
    
    $"x \in ON"$ abbreviates \("\)x is transitive and $\in$ well-order x(\")
    
    $" \forall x \in ON \varphi(x)"$ abbreviates $" \forall x (( x is transitive \wedge \in well-order x) \rightarrow \varphi(x))"$
    
    $"x \subseteq ON"$ , $"x \cap ON"$ are similarly abbreviated
    
    \textbf{\underline{Theorem}} $ON$ is well-ordered by $\in$, i.e. 
    \begin{enumerate}
        \item $\forall \alpha, \beta, \gamma \in ON ((\alpha < \beta \wedge \beta < \gamma) \rightarrow (\alpha < \gamma))$
        \item $\forall \alpha \in ON (\alpha \neg < \alpha)$
        \item $\forall \alpha, \beta \in ON ( \alpha < \beta \vee \beta < \alpha \vee \alpha = \beta)$
        \item $\forall z ((z \notin \emptyset \wedge z \subseteq ON) \rightarrow z has an \in - minimal element)$
        \end{enumerate}

    \underline{cor} There is no set which contains all ordinals.
    \underline{pf of cor} suppose X is a set which contains all ordinal
    By comprehension let $Y=\{Z \in X : Z is an ordinal\}$
    Then $Y=ON$ is a set.
    Theorem implies Y is transitive
    (1) If $alpha \in \beta$ and $\beta \in Y$ then $(\alpha < \beta)$ and $\alpha \in Y$ Y is well-ordered by $\in$
    Hence Y is an ordinal so $Y \in Y$
    but since Y is an ordinal by (2) of Theorem $Y \notin Y$ so contradiction

    \emph{Note} This is known as the Burali-Forti Paradox.

    we need some lemmas before proving the theorem

    \underline{Lem 1} ON is a transitive class, i.e. if $\alpha \in ON$ and $z \in \alpha$ then $z \in ON$

    \underline{pf} let $\alpha$ be an ordinal.
    Then $\alpha$ is transitive.
    Since $z \in \alpha$ then $z \subseteq \alpha$.
    $\in$ well-orders $\alpha$, so $\in$ well-orders $z$.
    To see $z$ is transitive: suppose $x \in y$ and $y \in z$
    so $y \in \alpha$ so $y \subseteq \alpha$ (since $\alpha$ is transitive and $z \subseteq \alpha$)
    so $x \in \alpha$
    hence $x,y,z \in \alpha$ and $x \in y$ and $y \in z$ so $x \in z$
    since $\in$ well-orders $\alpha$ and hence $\in$ is a transitive relation in $\alpha$
    which is what we wanted.
    Hence $z$ is transitive and well-ordered by $\in$ , so $z$ is an ordinal

    \emph{ordinals} \underline{feb 4}

    Recall: A set z is a transitive if whenever $y \in z$ then $y \subseteq$, i.e. if $x \in y$ and $y \in z$ then $x \in z$.
    A set $\alpha$ is an \underline{ordinal} if $\alpha$ is transitive and $\in$ well-orders $\alpha$.
    $\alpha < \beta \leftrightarrow \alpha \in \beta$ and $alpha \leq \beta \leftrightarrow \alpha < \beta \vee \alpha = \beta$

    \underline{Theorem} $ON$ is well-ordered by $\in$

    \underline{Lem 1} $ON$ is a transitive class, i.e. $\alpha \in ON$ and $z \in \alpha$ then $z \in ON$ where $ON$ is the class of ordinals. $\alpha$ is an ordinal (is what the in means)

    \underline{Lem 2} For $\alpha , \beta \in ON$ $\alpha \leq \beta $ iff $\alpha \subseteq \beta$.
    \begin{proof}

    $\alpha \leq \beta$ means $\alpha \in \beta \vee \alpha = \beta$
    if $\alpha = \beta$ immediately $\alpha \subseteq \beta$
    if $\alpha \in \beta$ then $\alpha \subseteq \beta$ since $\beta$ is transitive.

    suppose $\alpha \subseteq \beta$
    if $\alpha = \beta$ we are done, so suppose not.
    let $X = \beta \\ \alpha \neq \emptyset$ , $X \subseteq \beta$
    $\beta$ is wel-ordered by $\in$,
    X has an $\in$-minimal element $\xi$
    show $\xi = \alpha$
    For $\mu \in \xi$ we have $\mu \in \beta$ and $\mu \notin X$ so $\mu \in \alpha$
    Hence $\xi \leq \alpha$.
    Suppose $\xi \neq \alpha$.
    Then there is a$\mu \in \alpha \setminus \xi$
    $\mu$ and $\xi$ both in $\beta$.
    $\beta$ is totally ordered by $\in$ so one of $\mu \in \xi$, $\xi \in \mu$, or $\xi = \mu$.
    $\mu \neq \xi$ by choice of $\mu$
    since $\mu \in \xi$ and $\xi \notin \alpha$ $(\xi \in X = \beta \setminus \alpha)$ so $\mu = \xi$
    if $\xi \in \mu$ then $\xi \in \alpha$ (since $\mu \in \alpha$) but $\xi \notin \alpha$, hence $\xi \notin \mu$
    Hence $\xi = \alpha$, so $\alpha \in \beta$.
    \end{proof}

    \underline{Lem 3} If $\alpha, \beta \in ON$ then $\alpha \cap \beta \in ON$.
    (we will see $\alpha \cap \beta = \min(\alpha, \beta)$)

    \begin{proof}
        $\alpha \cap \beta \subseteq \alpha$ so $\alpha \cap \beta$ is well-ordered by $\in$.
        $\alpha \cap \beta$ is transitive since if $x \in \alpha \cap \beta$.
        then $x \in \alpha$ so $x \subseteq \alpha$, and $x \in \beta$ so $x \subseteq \beta$.
        Hence $x \subseteq \alpha \cap \beta$.
        So $\alpha \cap \beta$ is an ordinal.
    \end{proof}

Proof of Theorem:
\begin{proof}
    \begin{enumerate}
        \item $\forall \alpha, \beta, \gamma \in ON$ and $\alpha < \beta \wedge \beta < \gamma$ then $\alpha < \gamma$
        \begin{itemize}
            \item $\alpha$ is a transitive set.
            \item $\beta < \gamma$ means $\beta \in \gamma$.
            \item $\alpha < \beta$ means $\alpha \in \beta$.
            \item So $\alpha \in \gamma$ i.e. $\alpha < \gamma$.
        \end{itemize}
        \item $\forall \alpha \in ON$ $\alpha \notin \alpha$
        \begin{itemize}
            \item $\alpha$ is well-ordered by $\in$.
            \item If $\alpha \in \alpha$ then $\forall x \in \alpha$ ($x \notin x$).
            \item Hence $\alpha \notin \alpha$.
        \end{itemize}
        \item $\forall \alpha, \beta \in ON$ ($\alpha \in \beta \vee \beta \in \alpha \vee \alpha = \beta$)
        \begin{itemize}
            \item Let $\delta = \alpha \cap \beta$.
            \item $\delta \in ON$ by Lemma 3.
            \item $\delta \subseteq \alpha$ so $\delta \leq \alpha$ by Lemma 2.
            \item So either $\delta = \alpha$ or $\delta \in \alpha$.
            \item If $\delta = \alpha$ then $\alpha \subseteq \beta$ so either $\alpha = \beta$ or $\alpha \in \beta$ and we are done.
            \item So assume $\delta \in \alpha$
            \item $\delta \subseteq \beta$ so either $\delta = \beta$ or $\delta \in \beta$.
            \item if $\delta = \beta$ then $\beta \subseteq \alpha$ so we are done.
            \item so assume $\delta \in \beta$
            \item so $\delta \in \alpha \cap \beta = \delta$ but $\delta \notin \delta$, contradiction
            \item so either $\delta = \alpha$ or $\delta = \beta$ and we are done
        \end{itemize}
        \item $\forall z$ ($z \neq \emptyset \wedge z \subseteq ON \rightarrow$ $z$ has an $\in$-least element.
        \begin{itemize}
            \item Let $\alpha \in z$ be arbitrary.
            \item If $\alpha$ is $\in$-minimal in $z$ we are done.
            \item Suppose not.
            \item Consider $\alpha \cap z = \{\beta \in z : \beta < \alpha\}$
            \item $\alpha \cap z \subseteq \alpha$
            \item if $\alpha \cap z = \emptyset$ we are done as $\alpha = \emptyset$ would be $\in$-least in $z$
            \item if $\alpha \cap z \neq \emptyset$ then there is an $\in$-minimal element of $\alpha \cap z$, call this $\gamma$
            \item $\gamma \in \alpha \cap z$ and for any $\beta \in \alpha (\beta \notin \alpha \cap z)$
            \item if $\beta \in \gamma$ then $\beta \in \alpha$ hence $\beta \notin z$
            \item Hence $\gamma$ is $\in$-minimal element of $z$ which is what we wanted
        \end{itemize}
    \end{enumerate}
\end{proof}

    \underline{Ex}
    \begin{itemize}
        \item $0=\emptyset$ is the least ordinal
        \item $1 = \{\emptyset\}$ is the next biggest ordinal (no ordinal $\alpha$ with $0 < \alpha$ and $\alpha < 1$)
    \end{itemize}

    \emph{Note} If $\alpha \in ON$ then $\alpha = \{\gamma \in ON : \gamma < \alpha\}$

    Recall $S(x) = x \cup \{x\}$

    \underline{Lem} If $\alpha \in ON$ then $S(x) \in ON$.
    \begin{proof}
        If $\beta \in S(\alpha)$ then either $\beta \in \alpha$ or $\beta = \alpha$.
        In both cases $\beta \subseteq \alpha$ hence $\beta \subseteq S(\alpha)$.
        So $S(x)$ is transitive.
        If $z \subseteq S(x)$ and $z \neq \emptyset$.
        If $z = \{\alpha\}$ then $\alpha$ is an $\in$-minimal element of $z$ (since $\alpha \notin \alpha$)
        Otherwise, $z \cap \alpha \neq \emptyset$ has an $\in$-minimal element since $\alpha \in ON$ which will be an $\in$-minimal element of $z$.
    \end{proof}

    \underline{Lemma} For $\alpha \in ON$, $S(\alpha)$ is the immediate succesor of $\alpha$, i.e. $\alpha < S(x)$ and there is no ordinal $\beta$ with $\alpha < \beta$ and $\beta < S(\alpha)$
    \begin{proof}
        $\alpha \in S(\alpha) = \alpha \cup \{\alpha\}$ so $\alpha < S(\alpha)$
        If $\beta < S(alpha)$ then $\beta \in \alpha \cup \{alpha\}$
        So either $\beta \in \alpha$ or $\beta = \alpha$
        i.e. either $\beta < \alpha$ or $\beta = \alpha$ so can't have $\alpha < \beta$
    \end{proof}

    \textbf{Def} An ordinal $\beta$ is a \underline{successor ordinal} if $\beta = S(\alpha)$ for some $\alpha$.
    Otherwise $\beta$ is a \underline{limit ordinal} if $\beta \neq 0$ and $\beta$ is not a successor ordinal.
    A \underline{finite ordinal} or \underline{natural number} is if every $\alpha \subseteq \beta$ is either 0 or a successor ordinal.

    \emph{Note:} Axioms so far only produce finite ordinal e.g. 0,1,2,....

    \emph{Note:} $\beta \neq 0$ is a limit ordinal $\iff$ $\forall \alpha < \beta$ $S(\alpha) < \beta$

    \subsection{Axiom 7: Axiom of Infinity}
    $\exists X (\emptyset \in X \wedge \forall y (y \in X \rightarrow S(y) \in X))$

    \emph{Note: } We want to see that this X is an infinite set containing every finite ordinal.

    \underline{Lemma: } If $n$ is a natural number then $S(n)$ is a natural number.
    \begin{proof}
        $S(n)$ is an ordinal from earlier.
        If $\alpha \leq S(n)$ then
        If $\alpha = S(n)$ then $\alpha$ is a successor (done)
        Otherwise if $\alpha < S(n)$ then $\alpha \in n \cup \{n\}$
        So $\alpha \leq n$ so either $\alpha = 0$ or $\alpha$ is a successor.
    \end{proof}

    \underline{Theorem: } \underline{Principle of Ordinary Induction}
    For any set $X$, if $\emptyset \in X$ and $\forall y \in X (S(y) \in X)$ then $X$ contains all natural numbers.

    \begin{proof}
        Suppose the above is true.
        Suppose $n$ is a natural number with $n \notin X$.
        Let $Y = S(n) \setminus X \neq \emptyset$ since $n \in Y$.
        Y is a set of natural numbers.
        Y has an $\in$-minimal element, call it $k$.
        $k \leq n$
        Either $k=0$ or $k = S(j)$ for some $j$.
        $k \neq 0$ since $k \notin X$ and $0 \in X$
        $k \neq S(j)$ since $j \notin Y$ ($j \in k$ and $k \in$-minimal in $Y$)
        if $k = S(j)$ then $j \in S(n)$ so $j \in X$ so $S(j) = k \in X$, contradiction
        Hence no natural number $n \notin X$
    \end{proof}

    \textbf{Def} Let $\omega = \{n \in X: n is a natural number\}$ where $X$ is the set from the Axiom of Infinity.
    So $\omega$ is the set of natural numbers sometimes denoted $\mathbb{N}$

    \emph{Note:} we will see that $\omega$ is an ordinal.
    $\omega$ is the least infinite ordinal.
    $\omega$ is the least limit ordinal.

    \underline{\emph{Feb 6.}}
    Recall $\omega = \{n \in X : n is a natural number\}$ where $X$ is the set from the Axiom of Infinity so that $\emptyset \notin X \wedge \forall Y (Y \in X \rightarrow S(X) \in X)$

    \underline{Lemma: } Suppose $X$ is a set of ordinals which is an initial segment of $ON$. i.e. $\forall \beta \in X \forall \alpha \in \beta (\alpha \in X)$ then $X \in  ON$

    \begin{proof}
        $X$ is well-ordered by $\in$ since any set of ordinals is well-ordered by $\in$

        X is transitive since if $\alpha \in \beta$ and $\beta \in X$ then $\alpha < \beta$ so $\alpha \in X$

        Hence $X$ is an ordinal.

    \end{proof}

    \underline{Lemma: } $\omega$ is the least-limit ordinal.
    \begin{proof}
        $\omega$ is an ordinal since if $n \in \omega$ then $n$ is a natural number and if $m < n$ then m is a natural number, hence $m \in \omega$

        so $\omega$ is an initial segment of $ON$

        $\omega \neq \emptyset$

        suppose $\omega = S(y)$ for some $y$ then $y \in S(y) = \omega$ so $y$ is a natural number so $S(y) = \omega$ would be a natural number, so $\omega \in \omega$ contradicts that $\alpha \notin \alpha$ for any ordinal $\alpha$

        hence $\omega$ is a limit ordinal.

        every $\alpha < \omega$ is a natural number so either 0 or $S(n)$ some n

        hence $\omega$ is the least limit ordinal

    \end{proof}

    \emph{Note: } Now we have $0,1,2,\dots,\omega,S(\omega), S(S(\omega)),\dots$.
    Like the think of $S(\omega)$ as $\omega + 1$, etcetera.
    Need to define ordinal arithmetic $\alpha + \beta$, $\alpha \dot \beta$, $\alpha^{\beta}$.
    we will define these in terms of certain well-orderings we define.
    See that every well-ordering corresponds to a unique ordinal.

    \underline{Ex} subsets of $\mathbb{R}$ with the usual ordering <.
    Let $A = \{1-\frac{1}{n} : n \geq 1\}$
    $(A ; <) \cong (\omega,\in)$
    Let $B = \{ 1-\frac{1}{n}: n \geq 1\}\cup \{1\}$
    $(B;<) \cong (S(\omega); \in)$ where $S(\omega) = \omega + 1$
    Let $C = \{1-\frac{1}{n}:n\geq1\}\cup\{2-1/n:n\geq1\}$
    $(C;<) \cong (\omega + \omega;\in)$


    \underline{Lemma} If $\alpha$ and $\beta$ are ordinals and $f : \alpha \leftrightarrow \beta$ is an isomorphism from $(\alpha,\in)$ onto $(\beta,\in)$ then $\alpha = \beta$ and $f$ is the identity function.

    \begin{proof}
        Fix $\xi \in \alpha$, sp $f(\xi) \in \beta$.
        If $\zeta < \xi$ then $f(\zeta) < f(\xi)$ and vice versa.
        Then $\{\nu \in \beta : \nu < f(\xi)\}=\{f(\mu) : \mu \in \alpha \wedge \mu < \xi\}$.
        Hence $f(\xi) = f``\xi = \{f(\gamma) : \gamma \in \xi\}$
        Claim that $f(\xi) = \xi$ for all $\xi \in \alpha$.
        Let $X = \{\xi \in \alpha : f(\xi) \neq \xi\}$
        If $X = \emptyset$ we are done so suppose not.
        Let $X \neq \emptyset$ then $X$ is a set of ordinals so there is a least $\xi \in X$.
        If $\mu < \xi$ then $f(\mu)=\mu$.
        So $f(\xi)=\{f(\mu) : \mu < \xi\} = \{\mu : \mu < \xi\} = \xi$ contradicting that $f(\xi) \neq \xi$.
        Hence $f(\xi) = \xi $ for every $\xi \in \alpha$ i.e. $f$ is the identity map, and $\beta = \alpha$.
    \end{proof}

    \emph{Note: } There can be bijections between unequal ordinals, but they won't be order preserving.

    \underline{Ex: } $\omega = 0,1,2,\dots$ and $\omega + 1 = 0,1,2,\dots,\omega$.
    Pairing $\omega$ with 0 and the rest with their successors, so we have a bijection

    \emph{Note: } In the above proof we used the \underline{principle of transfinite induction}.
    Every non-empty set of ordinals has a least element.
    So fi $\varphi(x)$ is a formula and for all $\alpha \in ON$
    $(\forall \beta < \alpha \varphi(\beta)) \rightarrow \varphi(\alpha)$
    Then $\varphi(\alpha)$ holds for all ordinals $\alpha$

    \emph{Note: } Often split into cases for $\alpha$.
    show $\varphi(0)$.
    show $\varphi(\beta) \rightarrow \varphi(S(\beta))$.
    for a limit ordinal $\gamma$ show $(\forall \beta < \gamma \varphi(\beta)) \rightarrow \varphi(\gamma)$
    (orders are rigid we cant have an order isomorphism? maybe?)

    \underline{\textbf{Theorem}}: If R is a well-ordering of A then there is a unique ordinal $\alpha$ so that $(A;R) \cong (\alpha;\in)$
    (ordinal gives a representation of all orderings up to isomorphism)

    \begin{proof}
        $\alpha$ is unique from the previous lemma since if $f : (A;R) \leftrightarrow (\alpha;\in)$ and $g : (A;R) \leftrightarrow (\beta;\in)$.
        Then $g \circ f^{-1} : (\alpha;\in) \leftrightarrow (\beta;\in)$.
        So $g \circ f^{-1} = identity$ i.e. $f=g$ and $\alpha = \beta$
        For $a \in A$ let $a \downarrow = \{x\in A : x R a\}$.
        Then $a \downarrow$ is also well-ordered by R.
        Call $a \in A$ \underline{good} if $(a \downarrow;R) \cong (\xi; \in)$ for some ordinal $\xi$.
        Let $G = \{a \in A : a is good\}$.
        Define $f : G \rightarrow ON$ by setting $f(a) = $ the unique $\xi \in ON$ so that $(a\downarrow;R)\cong(\xi;\in)$
        Also for $a \in G$ let $h_a$ be the unique isomorphism $h_a=(a\downarrow;R)\cong(f(a);\in)$ ($f(a)=\xi$)
        This is justified by replacement
        If $a\in G$ and $c R a$ (i.e. $c \in a\downarrow$) then $c \in G$ since $c \downarrow \subseteq a\downarrow$
        $h_c = h_a \upharpoonright (c\downarrow): (c\downarrow;R)\cong(h_a(c);\in)$
        ($\upharpoonright$ is 'restricted')
        (Lecture drawing 54 mins into class)
        So $f(c) = h_a(c) < f(a)$
        Hence $f : G \rightarrow ON$ is order-preserving, i.e. $cRa \leftrightarrow f(c) < f(a)$.
        $ran(f)$ is an initial segment of $ON$ since if $\xi \in  ran(f)$ there is some $a \in G$ with $f(a) = \xi$
        So if $\mu < \xi`$ there is $c \in a\downarrow$ with $h_a(c) = \mu$
        But then $c \in G$ and $f(c) = h_a(c) = \mu$ so $\mu \in ran(f)$
        Hence $ran(f)$ is an ordinal $\alpha$ and f is an isomorphism from $(G;R)$ onto $(\alpha;\in)$.
        Want to show $G=A$.
        Suppose not, then let $e$ be the R-least elements of $A \setminus G$
        Then $e \downarrow = G$ since G is an R-initial segment of A.
        Hence $(e \downarrow;R) = (G;R) \cong (\alpha;\in)$.
        Hence $e \in G$, contradiction.
        Hence $G=A$ so $(A;R)\cong(\alpha;\in)$
    \end{proof}

    \textbf{Def: } For a well-ordering $(A;R)$ the \underline{order type} of $(A;R)$ , $type(A;R)$ is the unique ordinal $\alpha$ such that (A;R)$\cong$($\alpha;\in$)

    Recall: For (A;R) and (B;s) the \underline{lexicographic product} is the ordering of $A \times B$ given by $(a_1,b_1) <_{lex}(a_2,b_2) \leftrightarrow a_1 R a_2 \vee (a_1 = a_2 \wedge b_1 s b_2)$.

    \textbf{Def: } \underline{Ordinal arithmetic}
    For $\alpha,\beta \in ON$
    $\alpha \dot \beta = type(\beta \times \alpha, <_{lex})$
    $\alpha + \beta = type(\{0\}\times \alpha \cup \{1\} \times \beta, <_{lex})$

    \underline{Ex: } $\omega + 1$: $\{0\} \times \omega \cup \{1\}\times 1$ (see drawing at feb 6. 72 mins)

    Recall \underline{Ordinal Arithmetic}
    - $\alpha \dot \beta = type(\beta \times \alpha, <_{lex})$
    - $\alpha + \beta = type(\set{0} \times X \cup..??)$

    \underline{Ex: }
    $\omega \dot \omega = \langle 0,0 \rangle < \langle 1 , 0 \rangle < ....$

    multiplication and addition not commutative

    $\omega \dot 2 = \omega + \omega$
    $2 \dot \omega = 2 + 2 + ... = \omega$

    \emph{Note: } We can give recursive definitions.
    \textbf{Def: }
    - $\alpha + 0 = \alpha$
    - $\alpha + S(\beta) = S(\alpha + \beta)$
    - $\alpha + \lambda = Sup_{\beta < x} \alpha + \beta, \lambda limit$

    \textbf{Def: }
    - $\alpha \dot 0 = 0$
    - $\alpha \dot S(\beta) = \alpha \dot \beta + \alpha$
    - $\alpha \dot 2 = Sup(\beta < \lambda) (\alpha \dot \beta)$

    \textbf{Def: } \underline{Ordinal Exponentiation}
    - $\alpha^0 = 1$
    - $\alpha^{S(\beta)} = \alpha^\beta \dot \alpha$
    - $\alpha^\lambda = Sup(\beta < \lambda) \alpha^\beta$

    \underline{Ex: }
    - $\alpha^1 = 1 \dot \alpha = \alpha$
    - $\omega^2 = \omega \dot \omega$

    \subsection{Properties of Ordinal Arithmetic}
    \underline{Lemma}: addition and multiplicaiton are associative (distribute only on the left)
    - $(\alpha + \beta) + \gamma = \alpha + (\beta + \gamma)$
    - $(\alpha \dot \beta) \dot \gamma = \alpha \dot (\beta \dot \gamma)$
    - $\alpha \dot (\beta + \gamma) = \alpha \dot \beta + \alpha \dot \gamma$
    - $\alpha + \beta = \alpha + \gamma \rightarrow \beta = \gamma$
    - $\alpha \leq \beta \rightarrow \exists! \gamma (\alpha + \gamma = \beta)$
    - $\alpha > 0 \rightarrow \exists! \gamma, \delta \beta= \alpha \dot \gamma + \delta \wedge \delta < \alpha)$
    - $\alpha^{\beta + \alpha}$.... etc. seee feb 11. 15 mins

    But:
    $\omega + 1 \neq 1 + \omega = \omega$
    $\omega \dot 2 = \omega + \omega \neq 2 \dot \omega = \omega$
    $(1+1) \dot \omega = 2 \dot \omega = \omega \neq 1 \dot \omega + 1 \dot \omega = \omega + \omega$
    $1 + \omega = \omega = 2 + \omega$ but $1 \neq 2$

    \emph{Note:} We can prove these i n a couple ways.
    Transfinite induction (later).
    Comparing Order Types.

    \underline{Ex: } Show multiplicaiton is associative:
    $ (\alpha \dot \beta) \dot \gamma = \alpha \dot (\beta \dot \gamma)$
    \begin{proof}
        Find order-preserving bijection between two orderings.
        $\alpha \dot \beta = type(\beta \times \alpha, <_{lex})$
        $(\alpha \dot \beta) \dot \gamma = type(\gamma \times (\beta \times \alpha), <_{lex})$
        $a_1, a_2 \in \alpha b_1, b_2 \in \beta c_1,c_2 \in \gamma$
        $\langle c_1 \langle b_1, a_1 \rangle \rangle R < \langle c_2, \langle b_2,a_2 \rangle \rangle$
        $\iff c_1 < c_2 \vee ( c_1 = c_2 \wedge \langle \langle b_1, a_1 \rangle <_{lex} \langle b_2, a_2 \rangle ))$
        $iff c_1 < c_2 \vee ( c_1 = c_2 \wedge (b_1 < b_2 \vee (b_1 = b_2 \wedge a_1 < a_2)))$
        R corresponds to $(\alpha \dot \beta) \dot \gamma$
        Now $\alpha \dot (\beta \dot \gamma) = type ((\beta \times \gamma, <_{lex})\times \alpha, <_{lex})$
        $\langle \langle c_1,b_1\rangle,a_1\rangle S \langle\langle c_2,b_2,\rangle a_2\rangle$
        $iff (\langle c_1,b_1\rangle <_{lex})$ \dots etcetera

        (*) $\langle c_1,\langle b_1, a_1 \rangle \rangle R \langle c_2, \langle b_2, a_2 \rangle \rangle$
        $\iff \langle \langle c_1, b_1 \rangle , a_1 \rangle S \langle \langle c_2, b_2 \rangle , a_2 \rangle$

        Let $f : \gamma \times (\beta \times \alpha) \rightarrow (\gamma \times \beta) \times \alpha$
        be defined by:
        $f(\langle c, \langle b, a \rangle \rangle) = \langle \langle c, b \rangle , a \rangle$
        f is a bijection.
        f is order-preserving i.e. * from above
        shows $\langle c_1, \langle b_1, a_1 \rangle \rangle R \langle c_2, \langle b_2, a_2 \rangle \rangle$
        $\iff f(\langle c_1, \langle b_1, a_1 \rangle \rangle) S f(\langle c_2, \langle b_2, a_2 \rangle \rangle )$
    \end{proof}

    \subsection{Induction and Recursion on the Ordinals}
    \underline{induction:} used to prove statements.
    \underline{recursion:} used to define functions.

    \underline{Ex: } (natural numbers).
    Prove with induction that $\sum{i=1}{n} i = \frac{n(n+1)}{2}$
    $n=0 0=0$
    assume $\sum{i=1}{n} = \frac{n(n+1)}{2}$
    show $\sum{i=1}{n+1} = \frac{(n+1)(n+2)}{2}$

    define $n!$ by recursions
    $0! = 1$
    $(n+1)! = (n+1) \dot n!$

    \underline{\textbf{Theorem} Transfinite Induction on ON}
    For each formula $\varphi (x)$ if $\varphi(\alpha)$ holds for some ordinal $\alpha$, then there is a least ordinal $\xi$ so that $\varphi(\xi)$ holds

    \emph{Note: } We can replace `` holds `` by `` fails ``
    \begin{proof}
        Fix $\alpha$ sp that $\varphi(\alpha)$ holds
        Let $x = \set{\xi \in \alpha : \varphi(\xi)}$
        If $ x = \emptyset$ then $\alpha$ is the least ordinal with $\varphi(\alpha)$
        If $x \neq \emptyset$ then $x$ has an $\in$-least elements $\xi$, which is the least ordinal with $\varphi(\xi)$
    \end{proof}

    \underline{Corr} Suppose $\varphi(\xi)$ satisfies:
    (*) $\forall a \in ON (( \forall \beta < \alpha \varphi(\beta)) \rightarrow \varphi(\alpha))$
    Then $\varphi(\alpha)$ holds for every $\alpha \in ON$

    \begin{proof}
        If $\varphi(\alpha)$ fails for some $\alpha$ then there is a least $\alpha$ so $\varphi(\alpha)$ fails.
        Then $\forall \beta < \varphi(\alpha)$
        So by (*) we have $\varphi(\alpha)$ holds, a contradiction
    \end{proof}

    \emph{Note: } Odten we split (*) into cases
    $\alpha = 0$
    $\alpha = S(\beta)$ show if $\varphi(\alpha)$ holds then $\varphi(S(\beta))$ holds.
    $\alpha = \lambda$ is a limite: Show if $\varphi(\beta)$ holds for all $\beta < \alpha$ then $\varphi(\alpha)$ holds.

    \underline{Ex: } Show for every $\alpha \in ON$
    There are unique $\lambda$ and $n$ so that either $\lambda = 0$ or $\lambda$ is a limit ordinal.
    n is a natural number, and $\alpha = \lambda + n$
    prove by transfinite induction
    \begin{proof}
        base case:
        $\alpha = 0 = 0 + 0$
        So $\lambda = 0$ and $n = 0$
        case:
        $\alpha = S(\beta)$ Assume $\beta = \lambda + n$ some unique $\lambda$ and $n$
        $S(\beta) = \lambda + n + 1 = \lambda + S(n)$, $S(n)$ is a natural number.
        case:
        $\alpha = \lambda$ is a limit ordinal
        $\alpha = \lambda + 0$ 0 is a natural number.
    \end{proof}

    \underline{Transfinite Recursion on ON}
    We want to define funcitons on ON.
    we mean to find a formula $\varphi(x,y)$ so that $\forall x \exists! y \varphi(x,y)$.
    We can define $G(x) = !y \varphi(x,y)$ (not literallt a function)
    For any ordinal $\xi$ $G \upharpoonright \xi$ is a function (G restricted to xi)

    \underline{\textbf{Theorem} Primitive Recursion on ON}
    Suppose $\varphi(x,y)$ is a formula so that $\forall x \exists! y \varphi(x,y)$.
    Let $G(s) = $ the unique $y$ with $\varphi(s,y)$.
    We can define a formula $\psi(x,y)$ so that
    (1) $\forall x \exists! y \psi(x,y)$
    (2) $\forall \xi \in ON F(\xi) = G(F \upharpoonright \xi)$ where $F(\xi)$ is the unique $y$ with $\psi(\xi, y)$

    \underline{Ex: } Fix an ordinal $\alpha \in ON$
    Define $E_\alpha(\beta) = \alpha^\beta$
    Let $\varphi(x,y)$ be :
    ($x$ is not a function with $dom(x) \in ON \wedge y = 0 ) \vee$
    ($x$ is a function with $dom(x) = \xi \in ON \wedge$
    $((\xi = 0 \wedge y=1)) \vee (\exists \beta (\xi = S(\beta) \wedge y = x(\beta) \dot \alpha) \vee (\xi is a limit ordinal \wedge y = \set{x(\beta) : \beta < \xi})))$ (?)

    Then there is a formula $\psi(x,y)$ so that setting $F(\xi) = G(F \upharpoonright \xi)$ where $G(x) = !y$ with $\varphi(x,y)$ (x is a function)
    Then $F(\xi) = \alpha^\xi$ as we defined earlier $E_\alpha(\xi)$

    \emph{Note: } $F \upharpoonright \xi$ will be a function and equals $\set{\langle \beta ,F(\beta) \rangle : \beta < \xi}$ a function with domain $\xi$


    \section{Feb 13}
    \underline{\textbf{Theorom:} Primitive Recursion on ON}
    (See slides)
    \begin{proof}
        For $\delta \in ON$ let $App(\delta,h)$ be the formula that says that $h$ is a $\delta$-approximation to $F$, i.e.
        $h$ is a funciton
        $dom(h)=\delta$
        $h(\xi)=G(h \upharpoonright \xi)$ for all $\xi < \delta$
        where $G(s)$ is the unique $y$ with $\varphi(s,y)$

        Let $\psi (x,y)$ be: $(x \notin ON \wedge y = \emptyset) \vee (x \in ON \wedge \exists \delta > X \exists h(App(\delta,h) \wedge h(x) = y))$.
        (We need to see that there are enough of the approximiation and that if we have diff approxes that go up to different things, we need ot see that hey agree?
        that for every x there is a unique y so that psi of x,y.)

        Need to see that this $\psi$ works.
        Verify $\forall x \exists! y \psi(x,y)$.

        \underline(Show two claims:)
        \underline{Claim(unique)}: $(\delta \leq \delta` \wedge App(\delta,h) \wedge App(\delta`,h`)) \rightarrow h=h` \upharpoonright \delta$
        \underline{Claim(Exist)}: $\forall \delta \exists h App(\delta,h)$

        Granting these two claims:
        We have $\forall x \exists! y \psi(x,y)$.
        Given $\xi \in ON$ let $\delta > \xi$ and let $h$ be the unique function with $App(\delta,h)$.
        Then $F \upharpoonright \delta = h$ so $F \upharpoonright \xi = h \upharpoonright \xi$.
        So $F(\xi) = h(\xi) = G(h \upharpoonright \xi) = G(F \upharpoonright \xi)$.
        as wanted so we need to verify these two claims.

        \underline{Proof of claim(u)} using transfinite induction
        Assuming $\delta \leq \delta` \wedge App(\delta,h) \wedge App(\delta`,h`)$ that $h(\xi)=h`(\xi) \forall \xi < \delta$
        (doing this by induction on $\xi$)
        Suppose not (there cant be a least failure)
        Let $\xi$ be the least element of $\set{\alpha < \delta : h(a) \neq h`(\alpha)}$ (want this to be empty)
        Then $h \upharpoonright \xi = h` \upharpoonright \xi$.
        Hence $h(\xi) = G(h \upharpoonright \xi) = G(h` \upharpoonright \xi) = h`(\xi)$
        since $App(\delta,h)$ and $App(\delta`,h`)$
        contradiction $\xi$ satisfied $h(\xi) \neq h`(\xi)$

        \underline{Proof of claim(e)} (also by transfinit induction)
        Show $\forall \delta \exists h App(\delta,h)$
        Suppose not and let $\delta \in ON$ be least so that $\neg \exists h App(\delta,h)$

        Three Cases:
        $\delta = 0$ is impossible since $App(0,\emptyset)$

        $\delta = S(\beta)$ for some $\beta$.
        There is $g_{\beta}$ so that $App(\beta,g_{\beta})$.
        Let $f = g_{\beta} \cup \set{\langle \beta, G(g_{\beta})}$
        Notice $g_{\beta} = f \upharpoonright \beta$ so $f(\beta) = G(g_{\beta}) = G(f \upharpoonright \beta)$.
        So $f(\xi) = G(f \upharpoonright \xi)$ for all $\xi < S(\beta) = \delta$
        Hence $App(\delta,f)$ holds, contradicting that $\neg \exists h App(\delta,h)$

        $\delta$ is a limit ordinal.
        For all $\beta < \delta$ there is a unique $g_{\beta}$ with $App(\beta,g_{\beta})$ since claim(u) holds.
        Let $f = \bigcup \set{g_{\beta} : \beta < \delta}$
        Because of claim(u), $f$ is a function. (unique valued determined)
        $dom(f) = \delta$
        Since for every $\xi < \delta$ there is a $\beta$ with $\xi < \beta < \delta$ ($\delta$ is a limit ordinal)
        $f(\xi) = g_{\beta}(\xi) = G(g_{\beta} \wedge \xi) = G(f \upharpoonright \xi)$.
        Hence $App(\delta,f)$ contradiction.

    \end{proof}


    \textbf{Def:} For $\alpha \in ON$ an \underline{$\alpha$-sequence} is a fucntion with domain $\alpha$.
    Write $S_{\beta}$ for $S(\beta)$ for $\beta < \alpha$ and $S$ is an $\alpha$-sequence.

    \underline{Ex:} $\alpha = \omega$ $S_0,S_1,S_2, \dots$

    \textbf{Def:} Let $\bigcup^0 x = x$
    $\bigcup^{n+1} x = \bigcup \bigcup^n x$ for $n \in  \omega$
    The \underline{transitive closure} of $x$, $trcl(x)$ or $TC(x)$, as
    $TC(x) = \bigcup \set{\bigcup^n x : n \in \omega}$

    \underline{Ex:} $ x = \set{\set{\emptyset,}\set{\set{\emptyset}}}$
    $\bigcup^0 x = x$
    $\bigcup^1 x = \set{\emptyset,\set{\set{\emptyset}}}$
    $\bigcup^2 x = \emptyset \cup \set{\set{\emptyset}} = \set{\set{\emptyset}}$
    $\bigcup^3 x = \set{\emptyset}$
    $\bigcup^4 x = \emptyset$
    $\bigcup^n x = \emptyset$ for all $n > 4$
    $TC(x) = \set{\emptyset , \set{\set{\emptyset}},\set{\emptyset}}$

    \underline{Lemma:}
    (1) $x \subseteq TC(x)$
    (2) $TC(x)$ is a transitive set
    (3) $TC(x)$ is the smallest transitive set $t$ with $x \leq t$
    (4) If $y \in x$ then $TC(y) \subseteq TC(x)$

    \begin{proof}
    (1) True since $x = \bigcup^0 x \subseteq TC(x)$
    (2) If $y \in TC(x)$ then there is $n \in \omega$ with $y \in \bigcup^n x$ so $y \subseteq \bigcup \bigcup^n x$
        So $y \subseteq \bigcup ^n+1 x \subseteq TC(x)$.
        So $TC(X)$ is transitive.
    (3) Suppose $x \subseteq t$ and $t$ is a transitive set.
        Suppose for contradiction $TC(x) \nsubseteq t$.
        Let $n$ be least so that $\bigcup^n x \nsubseteq t$.
        $n \neq 0$ since $x \subseteq t$ $x=\bigcup^0 x$.
        So $n=k+1$ for some $k$.
        if $y \in \bigcup^n x$ then $y \in \bigcup^{k+1} x = \bigcup \bigcup^k x$.
        So there is $z \in \bigcup^k x$ with $y \in z$.
        $\bigcup^k x \subseteq t$ since $n$ was least.
        So $y \in z \in \bigcup^k x \subseteq t$.
        $t$ is transitive so $y \in t$.
        Hence $\bigcup^n x \subseteq t$ contradiction.
        Hence $t$ is the smallest transitive set with $x \subseteq t$.
    (4) Holds since if $y \in x$ then $\bigcup^n y \subseteq \bigcup^{n+1} x$ for all $n \in \omega$.
        So $TC(y) \subseteq TC(x)$.
    \end{proof}


    \subsection{Power Sets}
    Want to see that we can collect all subsets of a given set.
    This will allow us to build "bigger" sets in terms of cardinality.

    \textbf{Axiom 8: Power Set Axiom}
    $\forall x \exists y \forall z (z \subseteq x \rightarrow z \in y)$

    \textbf{Def:} The \underline{power set} of a set $x$ is $\mathcal{P}(x) = \set{z : z \subseteq x}$

    \underline{Ex:} $x = \set{0,1,2}$ then $\mathcal{P} = \set{\emptyset, \set{0}, \set{1}, \set{2}, \set{0,1}, \set{0,2}, \set{1,2}, \set{0,1,2}}$

    \emph{Note:} If x is a finite set with $n$ elements then $\mathcal{P}(x)$ has $2^n$ elements.

    \underline{Ex:} $\mathcal{P}(\omega)$ is the set of all sets of natural numbers.

    \emph{Note:} If $f : A \rightarrow B$ is a function then $f \subseteq A \times B$ so $f \in \mathcal{P}(A \times B)$.
    So we can make the following definitions.

    \textbf{Def:} $B^A$ or $prescript(A)B$ is the set of all functions $f$ with $dom(f) = A$ and $ran(f) \subseteq B$

    \emph{Note} $B^A \subseteq \mathcal{P}(A \times B)$

    \emph{Note: } When $A$ and $B$ are ordinals, this is not the same as ordinal exponentiation.
    In this case usually write $prescript(\beta)\alpha$ for functions $f : \beta \rightarrow \alpha$ different from the ordinal $\alpha^{\beta}$

    \underline{Ex: } Ordinal exponentiation $\omega^{\omega} = \omega \dot \omega \dot \omega \dots = Sup(n < \omega) \omega \dot \omega \dots \omega$ (n times)
    $prescript(\omega)\omega = $ all functions $f : \omega \rightarrow \omega$

    \textbf{Def:} When $\alpha$ is an ordinal,. $A^{<\alpha} = \bigcup(\xi < \alpha) A^{\xi}$
    this is the set of all sequences of elements of A of length $<\alpha$.

    \underline{Ex: } $A^{<\omega}$ is the set of all finite sequences of elements of A.

    \textbf{Def: } For $a_0,a_1,\dots,a_{m-1} \in A$ we write $(a_0,a_1,\dots,a_{m-1})$ for the sequence of length $m$ $S$ with $S \in A^m$ with $S(i) = a_i$ for $i<m$.

    \emph{Note: } Direct product can be justified using poer set axiom without needing replacement axiom.


    \section{Feb 18}

    \underline{Cardinality}

    \defn $X \preceq Y$ if there is an injection $f: X \rightarrow Y$.
    $X \approx Y$ if there is a bijection if $X \leftrightarrow Y$.
    $X \prec Y$ if $X \preceq Y$ but $Y \neq X$.

    \underline{Ex: } $\omega \approx \omega + 1 = \omega \cup \set{\omega}$
    $f(0) = \omega$
    $f(n) = n-1$ for $n \geq 1$

    \underline{Ex: } $\omega \approx \omega + \omega$
    $f(2n) = n$
    $f(2n +1) = \omega + n$ for $n \in \omega$

    \underline{Ex: } $\omega \times \omega \approx \omega$
    $f(n,m) = \frac{(n+m)(n+m+1)}{2} + n$
    they have the same cardinality

    (see graph in slides)

    \underline{Ex: } $\mathbb{P}(\mathbb{N}) \preceq \mathbb{R}$
    for $A \leq \mathbb{N} f(A) = \sum{n \in A} \frac{1}{3^n}$

    \underline{Ex: } $\mathbb{R} \preceq \mathbb{P}(\mathbb{N})$
    Let $\langle q_n : n \in \omega \rangle$ enumerate $\mathbb{Q}$
    Let $f(x) = \set{n \in \mathbb{N} : q_n < x}$
    If $x \neq y$ there is some $q_n$ with either $x < q_n < y$ or $y < q_n < x$
    So either $n \notin f(x) n \in f(y)$ or the opposite, so $f(x) \neq f(y)$

    \underline{\textbf{Question}} is there a bijection between $\mathbb{R}$ and $\mathbb{P}(\mathbb{N})$?

    \underline{Lemma: }
    (1) $\preceq$ is transitive
    (2) If $X \subseteq Y$ then $X \preceq Y$
    (3) $\approx$ is an equivalence relation

    \input{notes-part2}
    \input{notes-p3}
    \input{p4}
\input{logic}

\end{document}

